---
title: "7: Lab - Generalized Linear Models"
author: "Environmental Data Analytics | John Fay and Luana Lima | Developed by Kateri Salk"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Answer questions on M5/A5
2. Answer questions on M6 - GLMs
3. Practice more application GLM to real datasets

## Overview of the types of generalized linear models

Key terminology
- Response: variable we are trying to predict (Y, dependent or target)
- Independent variables: variable used to predict the response (X, predictor, feature)
- Record: vector of predictors and outcome value from an observation
- Intercept: predicted value when X = 0
- Regression coefficient: slope of the regression line
- Fitted values: estimates of Y obtained from the regression line (aka prediction)
- Residuals: difference between observed and fitted values (errors)
- Least squares: method used to find line that minimizes squared sum of residuals

>> Simple linear regression: one predictor and one response

General workflow
1. View data with a scatterplot of Y vs X
  - Can you see a trend, do you need to transform an axis (eg log transform)
2. Create a linear model
  - Find the best fit line (ordinary least sq method)
  - Assume residuals are normal, sensitive to outliers
  - Assumes causation (be very careful with this!)
3. Examine the model summary and plots
  - Summary provides information on intercept and slope, p values for coefficients,
    residual standard error, multiple R-squared (how the variability in variable Y is explained
    by the variability of X -- want this value to be 1), F-statistic
  - Residuals vs Fitted: compares variables that are fitted by the model and the values that
    were actually observed, do not want to see patterns which indicates bias
  - Normal Q-Q: shows that values follow a normal distribution, deviations at the beginning and 
    end are to be expected
  - Scale-Location: fitted values are compared to standardized residuals (again, you do not want
    to see a pattern)
  - Residuals vs Leverage: leverage is the impact that each datapoint has on the final model product,
    if the leverage is small the datapoint does not have a large impact on the model, high residual 
    and high leverage together indicates that a value should be removed from the analysis


>> Multiple linear regression: many independent variables to predict response

- Can find variables with correlation matrices
- Issues with overfitting: you may explain the observation more than once if you are using variables
  that already have a correlation
- Akaike's Information Criterion (AIC): this is used to determine which variables can be removed from
  the model that can imporove the model

General Workflow
1. Generate linear model 
2.
3. 

>> ANOVA: predicting Y from categorical variables

- Categorical variables but be categorized as factors (variable used to group data that is suspected
  to explain variability in another response variable)!
- Levels:
- ??
- Two way ANOVA requires you to understand interactive effects

Assumptions
- Populations are normally distributed
- 
- 

Running in R
- aov(data = "", Y ~ X)
- lm(formula = Y ~ X, data = "")
- Post Hoc test: if means are found not to be the same, which are different?
  - Turkey HSD: compares all pairwise combinations (same letters = same mean)
  - Visualization with box plots to view means

>> T-test: comparing Y to a set equality

- Are ozone levels below the threshold for "poor" AQI index?
- Equality needs to be in the null hypothesis, and alternative covers the rest of the space
- Testing for the normality with Shapiro-Wilkes, ____, _____
- p value tells you to reject null hypothesis if <0.05
- Two-sample T-test is used to compare if two variables have the same mean
  - Same as two way ANOVA

## Set up
```{r, message = FALSE}
library(tidyverse)
library(agricolae)   
library(here)
here()

EPAair <- read.csv(here("Data/Processed_KEY/EPAair_O3_PM25_NC1819_Processed.csv"), stringsAsFactors = TRUE)
# Set date to date format
EPAair$Date <- as.Date(EPAair$Date, format = "%Y-%m-%d")

Litter <- read.csv(here("Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"), stringsAsFactors = TRUE)
# Set date to date format
Litter$collectDate <- as.Date(Litter$collectDate , format = "%Y-%m-%d")

# Set theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```



## Visualization and interpretation challenge

Create three plots, each with appropriately formatted axes and legends. Choose a non-default color palette.

1. geom_density of ozone divided by year (distinguish between years by adding transparency to the geom_density layer).
2. geom_boxplot of ozone divided by year. Add letters representing a significant difference between 2018 and 2019 (hint: stat_summary). 
3. geom_violin of ozone divided by year, with the 0.5 quantile marked as a horizontal line. Add letters representing a significant difference between 2018 and 2019. 

```{r}
#Exercise 1:

EPAair$Year <- as.factor(EPAair$Year)

ggplot(EPAair) +
  geom_density(aes(Ozone, fill = Year), alpha = 0.5) + # Can also use as.Factor()
  labs(x = "Ozone AQI Value",
       y = "Density")

ggplot(EPAair, aes(Year, Ozone)) +
  geom_boxplot() +
  stat_summary(geom = "text", # Going to be adding text
               fun = max, # Add the value at the max (top)
               vjust = -1,
               size = 4,
               label = c("b", "a")) +
  labs(x = "",
       y = "Ozone AQI Value") +
  ylim(0, 150)


ggplot(EPAair, aes(Year, Ozone)) +
  geom_violin(draw_quantiles = 0.5) +
  stat_summary(geom = "text", # Going to be adding text
               fun = max, # Add the value at the max (top)
               vjust = -1,
               size = 4,
               label = c("b", "a")) +
  labs(x = "",
       y = "Ozone AQI Value") +
  ylim(0, 150)

```

## Linear Regression

Important components of the linear regression are the correlation and the R-squared value. The **correlation** is a number between -1 and 1, describing the relationship between the variables. Correlations close to -1 represent strong negative correlations, correlations close to zero represent weak correlations, and correlations close to 1 represent strong positive correlations. The **R-squared value** is the correlation squared, becoming a number between 0 and 1. The R-squared value describes the percent of variance accounted for by the explanatory variables. 

For the NTL-LTER dataset, can we predict PM2.5 from Ozone?

```{r}

#Exercise 2: Run a linear regression PM2.5 by Ozone. Find the p-value and R-squared value. 

air_quality <- lm(data = EPAair, PM2.5 ~ Ozone)
summary(air_quality)
  # Resulting p-value: 2.2e-16
  # Resulting R-squared value: 0.1316 (multiple) and 0.1314 (adjusted)

#Exercise 3: Build a scatterplot. Add a line and standard error for the linear regression. Add the regression equation to the plot

ggplot(EPAair, aes(Ozone, PM2.5)) +
  geom_point() +
  geom_smooth(method = "lm") +
  annotate(geom = "text",
           x = 50, y = 90,
           size = 4,
           label = "Ozone = 0.38*PM2.5 + 15.64")+
  ylim(0,100)

```


## AIC to select variables

What other variables can we add to improve model?

```{r}

#Exercise 4: Build correlation plots and identify more possible explanatory variables to add to the regression.
library(corrplot)
library(dplyr)

EPAair$Month <- as.numeric(EPAair$Month)
EPAair$Year <- as.numeric(EPAair$Year)
EPAair$Ozone <- as.numeric(EPAair$Ozone)

Air_corr_data <- 
  EPAair %>%
  select(c(mean_Lat:Ozone)) %>%
  na.omit()

Air_corr <- cor(Air_corr_data)
corrplot(Air_corr, method = "ellipse")
corrplot.mixed(Air_corr, upper = "ellipse")
  # Results: next best option after Ozone is mean_Lat

#Exercise 5: Choose a model by AIC in a Stepwise Algorithm. Do the results from AIC match the variables you selected on Exercise 4?




#Exercise 6: Run another regression using the variables selected on Exercise 6. Compare r-squared value with the one from Exercise 2. 

```



## Litter Exercise

```{r}
# Wrangle the data
Litter.Totals <- Litter %>%
  group_by(plotID, collectDate, nlcdClass) %>%
  summarise(dryMass = sum(dryMass))

# Format ANOVA as aov
Litter.Totals.anova <- aov(data = Litter.Totals, dryMass ~ plotID)
summary(Litter.Totals.anova)

# Extract groupings for pairwise relationships
Litter.Totals.groups <- HSD.test(Litter.Totals.anova, "plotID", group = TRUE)
Litter.Totals.groups$groups

Litter.Totals <- Litter.Totals %>%
  mutate( treatgroups = Litter.Totals.groups$groups[plotID,2])

# Graph the results
Litter.Totals.plot <- ggplot(Litter.Totals, aes(x = plotID, y = dryMass)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  stat_summary(geom = "text", fun = max, vjust = -1, size = 3.5,
               label = c("abc", "ab", "abcd", "bcd", "bcd", "a", 
                         "d", "abcd", "cd", "d", "abcd", "abcd")) +
  labs(x = "Plot ID", y = "Dry Mass (g)") +
  ylim(0, 85)
print(Litter.Totals.plot)

```

```{r}
#Exercise 7: Improve the plot

