---
title: "Lab 3a - Exercises"
author: "Environmental Data Analytics | John Fay and Luana Lima "
date: "Fall 2025"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## Set up the session
- Load packaged: tidyverse, lubridate, & here
- Import the `EPAair_PM25_NC2019_raw.csv` file
  - Set strings to be read in as factors
  - Set the "Site.ID" column to be read in as a factor
- Set the date column to be a date

```{r Setup}
#Import packages
library(tidyverse); library(lubridate); library(here)

#Check workspace
here()
pm25_file <- here("Data", "Raw", "EPAair_PM25_NC2019_raw.csv")

#Import data
pm25 <- read.csv(
  pm25_file,
  stringsAsFactors = TRUE,
  colClasses = c("Site.ID"='factor')
)

#Fix the date
pm25$Date <- mdy(pm25$Date)

```


## Explore properties of the dataframe
-   How many different sites are included in the data set?
-   What is the mean daily PM 2.5 concentration?
-   In what units of are PM 2.5 concentrations reported?

```{r Explore the data}
# Look at column headers
colnames(pm25)

# Compute the number of different sites
length(summary(pm25$Site.ID)) # Summarize the count in each site then find the length of the vector
unique(pm25$Site.ID) # Yields the different sites and how many levels there are

# Compute the mean and median PM 2.5 concentration
mean(pm25$Daily.Mean.PM2.5.Concentration)
median(pm25$Daily.Mean.PM2.5.Concentration)

# What are the units
unique(pm25$UNITS) # Check how many unique units there are
print(pm25$UNITS[1]) # All the units are the same, so just print the first entry 

# What is the date range (min and max dates)
min(pm25$Date) # Min date
max(pm25$Date) # Max date

```


## Plot distribution of values

-   Number of samples using each different source 
-   Distribution of daily mean concentration values
-   Distribution of daily mean concentration, by source*
  *   There are multiple ways to visualize this...

```{r Visualize the data}
library(ggplot2)

#Number of samples using each different source 
ggplot(pm25) +
  geom_bar(aes(x = Source))

#Distribution of daily mean concentration values
ggplot(pm25) +
  geom_histogram(aes(Daily.Mean.PM2.5.Concentration))

#Distribution of daily mean concentration, by source
ggplot(pm25) +
  geom_freqpoly(aes(x = Daily.Mean.PM2.5.Concentration, color = Source)) +
  scale_color_manual(values = c("grey50", "blue"))

```

